Ace: An Actively-Typed Language and Compilation Environment for High-Performance Computing

### 2013-04-24 ace/ace.tex
Updated ace abstract and intro

\begin{abstract}
Researchers developing languages and abstractions for high-performance computing must consider a number of design criteria, including performance, verifiability, portability and ease-of-use. Despite the deficiencies of today's tools and the availability of apparently superior options, end-users have been reluctant to adopt new language-based abstractions. We argue that this can be largely attributed to a failure to consider three critical criteria: continuity, extensibility and interoperability. This paper introduces Ace, a language that incorporates several novel mechanisms and makes principled design choices in order to satisfy this more comprehensive set of criteria. Ace builds upon existing standards in HPC, particularly Python and OpenCL, and it can be modularly extended with abstractions drawn from several previously disparate paradigms. The core innovation supporting this is a novel reification of types as first-class objects at compile-time, a refinement to the concept of active libraries that we call \emph{active types}. We validate our overall design with a case study of a simulation framework enabling the modular specification and efficient execution of ensembles of neural simulations across clusters of GPUs.

@@

Professional end-users today generally use dynamically-typed high-level languages like MATLAB, Python, R or Perl for tasks that are not performance-sensitive, such as small-scale data analysis and plotting \cite{nguyen2010survey}. For portions of their analyses where the performance overhead of dynamic type checking and automatic memory management is too high, they will typically call into code written in a statically-typed, low-level language, most commonly C or Fortran, that uses low-level parallel abstractions like pthreads and MPI \cite{4222616,basili2008understanding}. Unfortunately, these low-level languages and abstractions are notoriously difficult to use and code correctness cannot be automatically verified in general.

@@

Ace is a programming language intended to be used by professional end-users as well as researchers in areas related to high-performance computing and computational science. The primary contribution of this paper consists of describing several novel language mechanisms and novel combinations of existing mechanisms that Ace uses to cohesively address the design criteria above. We anticipate that many of these mechanisms and choices will also influence other languages in the future. 
Following a principled design methodology, we justify each design decision by appeal to this full set of design criteria and citations to relevant literature. These criteria and methodology constitute another contribution of this work that we hope  will lead to fewer \emph{ad hoc} designs and serve as a foundation for evaluating competing designs in HPC and related domains in the future.

We begin in Section \ref{highlevel} with a high-level overview of the Ace language design and the reasoning behind the choices that were made. In Section \ref{usage} we provide further details by way of concrete examples employing Ace for low-level GPU programming, based on an internalization of the OpenCL programming language as an Ace library. Ace can be used both as a standalone language via the \verb|acec| compiler and as an interactive compilation environment. This latter mode of use enables just-in-time specialization and direct invocation of Ace functions from within Python scripts, as well as deep integration between the \verb|numpy| library, the OpenCL host API and the Ace compiler.

In Section \ref{att}, we describe the extensible type system, which is based on the novel idea of reifying types as metalanguage objects. This mechanism, called \emph{active typing}, was used to implement the OpenCL library. 
The operations associated with \emph{active types} are given meaning by translation to a user-defined backend. We show in Section \ref{translation} how this allows us to give semantics to OpenCL constructs by translation to OpenCL itself, as well as to CUDA and C99, increasing the operational flexibility of the OpenCL standard. Then in Section \ref{usecases}, we  continue with several examples of higher-level parallel abstractions that can be implemented in Ace. In other systems, these have required deriving a new variant of an existing language or a new language, both strategies that violate at least two of the three new criteria above. In Ace, these exist within libraries and can safely be used within a single program.

To demonstrate that Ace can be used for realistic scientific workloads, we briefly describe in Section \ref{casestudy} a case study that used Ace as the foundation for a neurobiological simulation framework capable of orchestrating the execution of thousands of realizations of a large stochastic neural circuit model on clusters of GPUs. 
We conclude in Sections \ref{related} and \ref{discussion} with related work and a discussion of some limitations of Ace at the time of writing, as well as a discussion of future work building on both Ace itself as well as the concept of active typing central to Ace.

\section{The Design of Ace}\label{highlevel}
It is now widely-acknowledged that static type systems are fundamental tools for performance and verifiability, both critical criteria in HPC, so Ace is fundamentally a statically-typed language. 
To achieve continuity, Ace is implemented as a library in the ubiquitous scripting language Python. The Ace grammar is precisely that of Python, and it share's Python's packaging system, so any tools that work with Python files and packages can immediately be used with Ace.

@@

\section{Usage of Ace}\label{usage}

\section{Active Typechecking and Translation}\label{att}

@@

>>>>

\begin{abstract}
Researchers developing languages and abstractions for high-performance computing must consider a number of design criteria, including performance, verifiability, portability and ease-of-use. Despite the deficiencies of legacy tools and the availability of seemingly superior options, end-users have been reluctant to adopt new language-based abstractions. We argue that this can be largely attributed to a failure to consider three additional criteria: continuity, extensibility\- and interoperability. This paper introduces Ace, a language that aims to satisfy this more comprehensive set of design criteria. To do so, Ace introduces several novel compile-time mechanisms, makes principled design choices, and builds upon existing standards in HPC, particularly Python and OpenCL. OpenCL support is not built into the language, but rather implemented atop an extensibility mechanism that also admits abstractions drawn from several other seemingly disparate paradigms. The core innovation underlying this and several other features of Ace is a novel reification of types as first-class objects at compile-time, a refinement to the concept of active libraries that we call \emph{active types}. We validate our overall design by considering a case study of a simulation framework enabling the modular specification and efficient execution of ensembles of neural simulations across clusters of GPUs.

@@

Professional end-users today generally use dynamically-typed high-level languages like MATLAB, Python, R or Perl for tasks that are not performance-sensitive, such as small-scale data analysis and plotting \cite{nguyen2010survey}. For portions of their analyses where the performance overhead of dynamic type checking and automatic memory management is too high, they will typically call into code written in a statically-typed, low-level language, most commonly C or Fortran, that uses low-level parallel abstractions like pthreads and MPI \cite{4222616,basili2008understanding}. Unfortunately, these low-level languages and abstractions are notoriously difficult to use and automatic verification is intractable in general.

@@

In this paper, we introduce Ace, a programming language targeting professional end-users as well as researchers across high-performance computing and related domains. Ace has followed a principled design methodology guided by this more comprehensive set of design criteria in order to avoid many of the issues that have hindered previous language-based approaches in high-performance computing. These criteria, this design methodology, and the novel mechanisms and designs developed to satisfy these criteria constitute the generalizable contributions of this paper. We also hope that Ace itself will be useful to the HPC community\footnote{\small Ace is openly available at http://acelang.org/.}, and we describe several use cases and an initial case study as a means to preliminarily validate it's utility.

 We begin in Section \ref{usage} with simple examples that show how Ace can be used for low-level GPU programming and introduce the fundamental decisions made in its design. We discuss the motivations, based on the criteria above, behind key design decisions, including the use of static typing, a fixed syntax based on Python, a novel type propagation and type inference scheme, and an explicit phase separation between compile-time and run-time logic.

The examples in Section \ref{usage} are based on an internalization of the OpenCL kernel programming language as a library. This library is built atop a novel extensibility mechanism that we call \emph{active typechecking and translation} (AT\&T) and detail in Section \ref{att}. This mechanism relies on the core idea of representing types as metalanguage objects. We refer to such types as \emph{active types} because they are a refinement to the concept of \emph{active libraries} (cf. Section \ref{related}). Unlike prior approaches to extensibility that take a language-global view, hindering  interoperability, the AT\&T mechanism guarantees that extensions are composable by construction.

%Despite its simplicity and safety, this mechanism is flexible enough to define the entirety of the OpenCL kernel language (which is derived directly from C99) and, combined with Ace's type propagation scheme, substantially enhance it's ease-of-use. 
To further demonstrate the flexibility of this mechanism, we continue in Section \ref{usecases} by outlining examples of other, higher-level parallel abstractions, including global address spaces, message passing and functional data parallelism, that can be cleanly implemented using AT\&T. First-class support for each of these has required a new language in the past. AT\&T enables compile-time logic to be coupled with user-defined types, and thus all of these first-class abstractions can safely be used within a single program or workflow.

 Ace can be used as a standalone language via the \verb|acec| compiler and also as an interactive compilation environment. This latter mode of use supports just-in-time specialization and direct invocation of Ace functions from within Python scripts, as well as deep integration between the widely-used \verb|numpy| library (including its internal type system),  OpenCL's host API (by a mechanism also available to other host APIs, such as CUDA's) and the Ace compiler itself. We describe this in Section \ref{compenv}. Achieving this deep integration makes further use of Ace's representation of types.

To demonstrate the utility of Ace for realistic scientific workloads, we briefly describe in Section \ref{casestudy} a case study that uses Ace as a compilation environment supporting a scientific simulation framework capable of orchestrating the execution of thousands of realizations of a large stochastic neurobiological circuit model on clusters of GPUs.

We conclude in Sections \ref{related} and \ref{discussion} with related work and a discussion of some limitations of Ace at the time of writing, as well as a discussion of future work that builds upon the concept of active typing central to Ace.

% a programming language intended to be used by professional end-users as well as researchers in areas related to high-performance computing and computational science. The primary contribution of this paper consists of describing several novel language mechanisms and novel combinations of existing mechanisms that Ace uses to cohesively address the design criteria above. We anticipate that many of these mechanisms and choices will also influence other languages in the future. 
%Following a principled design methodology, we justify each design decision by appeal to this full set of design criteria and citations to relevant literature. These criteria and methodology constitute another contribution of this work that we hope  will lead to fewer \emph{ad hoc} designs and serve as a foundation for evaluating competing designs in HPC and related domains in the future.

\section{Design and Usage}\label{usage}
It is now widely-acknowledged that static type systems are fundamental tools for performance and verifiability, both critical criteria in HPC, so Ace is fundamentally a statically-typed language. 
To achieve continuity, Ace is implemented as a library in the ubiquitous scripting language Python. The Ace grammar is precisely that of Python, and it share's Python's packaging system, so any tools that work with Python files and packages can immediately be used with Ace.

@@

\section{Active Typechecking and\\Translation (AT\&T)}\label{att}

@@

\section{Ace: A Compilation Environment}\label{compenv}
